---
title: "Produce a GTFS"
output: rmarkdown::html_vignette
abstract: "This vignette demonstrates producing GTFS using the croquis() Shiny app and functions including gtfs_to_ssfs() and ssfs_to_gtfs()"
vignette: >
  %\VignetteIndexEntry{Produce a GTFS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(croquis)
```

## 1. Introduction

**croquis** can be used to produce GTFS for planning and research purposes in minutes. This is possible thanks to the Simplified Speed and Frequency Specification (SSFS) data structure that **croquis** introduces.

You can produce a GTFS from scratch or by uploading an existing one, either through the `croquis()` Shiny App or by in-package functions including `gtfs_to_ssfs()` and `ssfs_to_gtfs()`. This vignette outlines the key ways you can produce a GTFS with **croquis**.

## 2. Producing a GTFS from scratch

To launch the Croquis Shiny App, run:

```r
croquis()
```

Using the app, you can follow the steps below to create a GTFS from scratch. This vignette also demonstrates how to use the other functions in **croquis** to build a simple GTFS in your R console without the Shiny App.

# 2.1. Getting started: setting agency details

On the Home page of the Shiny App, set the project location by searching for a city or setting the coordinates manually. This will center the map in the other modules to make network modification easier. Provide agency details by clicking `+ Add new agency` in the Agencies panel. Fill out these details

# 3. Producing a GTFS based on an existing 
