---
title: "SSFS data structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SSFS data structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(ggplot2)
library(croquis)
```

The Simplified Speed and Frequency Specification (SSFS) is the foundational data structure of Croquis. It consists of a simplified representation of a GTFS (General Transit Feed Specification) and contains the instructions for how to build one, which are read by `ssfs_to_gtfs()`.

Like the GTFS, it contains a list of tables pertaining to a transit network and schedule. In fact, three of the tables are shared (agency, routes, and calendar) and are simply passed on from one data structure to the other. Unlike GTFS, SSFS does not contain stop_times or trips. Instead, it contains:

- A simple features (geospatial) table called **itin** which contains the geometry of the distinct itineraries (or variants) of the routes.
- A data.frame table called **span** which contains the first and last departure time for each service window, for each service, and for each route itinerary. 
- A data.frame called **hsh** which details the headway (interval between trips) and speed by hour of operation for each route itinerary and service.
- A data.frame called **stop_seq** which details the stop order for each route itinerary as well as the speed factor to apply to adjust speeds by route segment relative to the speed provided by hour in *hsh*. This is used to produce stop_times in conjunction with the distance between stops based on the geometry in *itin*. 

SSFS was designed to facilitate production, editing and calibration of transit networks and schedules in a GTFS-compatible format. 

When you load a GTFS network into the Croquis Shiny app, it converts the GTFS to an SSFS. Note that many details of the GTFS are dropped in this conversion, in favour of a simpler data structure that facilitates rapid modification while retaining key details including how service levels and speeds vary throughout the day, and how speed varies across route segments.

To convert a GTFS to an SSFS in the console, use `gtfs_to_ssfs()`. To produce an empty SSFS in the console, use `ssfs()`.

The rest of this vignette covers each table of the SSFS, including the fields of each and data types. We will use `translink` data from Vancouver that is included in the croquis package.

### agency

The **agency** table consists of the required and conditionally required fields of the GTFS agency table. These are all `<chr>` fields. This table is passed on between GTFS and SSFS formats.

The `agency_id` field links this table to the *routes* table.

```{r, message = FALSE}
head(translink$agency)
```

### routes

The **routes** table consists of the required and conditionally required fields of the GTFS routes table, plus `route_color` and `route_text_color`. This table is passed on between GTFS and SSFS formats. If an imported GTFS is missing the colour-related fields or either of the conditionally required fields, `gtfs_to_ssfs()` will produce them.

Field connections:

- The `agency_id` field connects this table to the *agency* table.
- The `route_id` field connects this table to the *itin* table.

```{r, message = FALSE}
dplyr::glimpse(translink$routes)
```

### stops

**stops** in the SSFS is a simple features table taht consists of the stop_id and stop_name fields that are passed on between the SSFS and GTFS, plus a `POINT` geometry field based on the stop_lat and stop_lon of the GTFS.

The `stop_id` field connects this table to the *stop_seq* table.

```{r, message = FALSE}
dplyr::glimpse(translink$stops)
```

### itin

**itin** is a simple features table that identifies each unique route itinerary in the network and provides a `LINESTRING` geometry for each. Each itinerary corresponds to a unique trip pattern (combination of stop_id and stop_sequence) of a route. The geometry is based on data from `shapes` table of the GTFS. In the case of GTFS that does not include a `shapes` table, `gtfs_to_ssfs()` will create one by either connecting the stops directly or by connecting the stops along the road network using the `osrm` or `valh` packages, depending on associated route_type and the availability of OSRM or Valhalla routing servers.

*itin* also includes the `direction_id` and `trip_headsign` fields which are carried over from the GTFS trips table and are used to create trips data when producing a GTFS from the SSFS with `ssfs_to_gtfs()`.

Field connections:

- The `itin_id` field connects this table to the *stop_seq*, *span*, and *hsh* tables.
- The `route_id` field connects this table to the *routes* table.

```{r, message = FALSE}
dplyr::glimpse(translink$itin)
```

### calendar

The **calendar** table is carried over directly from the GTFS calendar table and consists of all of its required fields. It defines the operating dates of each service.

The `service_id` field connects to the *span* and *hsh* tables.

```{r, message = FALSE}
head(translink$calendar)
```

### span

The **span** table contains data on the first and last departure time for each service window, for each service, and for each route itinerary. A service window is a period during which continuous service is offered based on headways defined in the *hsh* table. 

Field connections:

- The `itin_id` field connects to the *itin*, *hsh*, and *stop_seq* tables.
- The `service_id` field connects to *calendar* and *hsh*.

```{r, message = FALSE}
dplyr::glimpse(translink$span)
```

### hsh

The **hsh** table details the headway (interval between trips) and speed by hour of operation for each route itinerary and service.

When importing a GTFS into Croquis or converting a GTFS to SSFS, `gtfs_to_ssfs()` calculates the average commercial speed of trips by route itinerary, service, and hour. This data is stored in the hsh table, to be used viewed and edited while in SSFS format, then converted into stop_times based on the hour-by-hour data.

```{r, message = FALSE}
dplyr::glimpse(translink$hsh)
```

As mentioned, the SSFS allows us to view and modify how speeds and service levels vary throughout the day. Converting a GTFS to SSFS can be useful for gathering summary information about service levels and speeds. Below is an example for generating a visual for daily service level and speed variation for weekday service on the R4 eastbound in Vancouver.

```{r, message = FALSE}
# Fetch route ID for R4
r4_route_id <- 
  translink$routes |> 
  filter(route_short_name=="R4") |> 
  pull(route_id)

# Fetch itin ID for R4 bound for Joyce-Collingwood Station
r4_e_itin_id <- 
  translink$itin |> 
  filter(route_id==r4_route_id,stringr::str_detect(trip_headsign,"Joyce")) |> 
  pull(itin_id)

# Fetch headway and speeds by hour table for weekday service and R4 bound for Joyce-Collingwood
r4_trips_ph_speed <- 
translink$hsh |> 
  filter(itin_id==r4_e_itin_id,service_id=="mon-fri") |> 
  mutate(
    trips_per_hour=floor(60/headway),
    hour_dep = as.numeric(stringr::str_sub(hour_dep,1,2))) |> 
  select(hour_dep,trips_per_hour,speed)

# Visualise trips per hour and speeds in a ggplot
ggplot2::ggplot(r4_trips_ph_speed,aes(x = hour_dep))+
  geom_col(aes(y = trips_per_hour), width = 0.8, alpha = 0.5)+
  geom_line(aes(y = speed)) +
  scale_y_continuous(
    name = "Columns: Departures per hour | Line: Speed (km/h)")+
  scale_x_continuous(breaks = r4_trips_ph_speed$hour_dep)+
  labs(
    title = "R4 Eastbound: Departures per hour and speed by hour (weekday)",
    x = "Hour of departure") +
  theme_minimal()
```

### stop_seq

The **stop_seq** table details the stop order for each route itinerary as well as the speed factor to apply to adjust speeds by route segment. When converting a GTFS to SSFS, speed_factors are calculated once per itinerary and per stop based on how the average interstop speed between that stop and the next one compares to the average speed of all trips of that itinerary. Speed factors are therefore not provided for the last stop in any stop sequence. 

```{r, message = FALSE}
dplyr::glimpse(translink$stop_seq)
```

The singular speed factor per stop is a heuristic approach that does not take into account how the relative speed of segments could vary throughout the day or depending on the service (weekday vs. weekend).

To demonstrate how the speed_factor field adjusts the speeds of a model to better represent the behaviour of a transit vehicle on a given route, below is a graph of the modelled speeds on the R4 route eastbound from UBC to Joyce-Collingwood for departures between 8:00AM and 8:59AM. Reflecting the data in the GTFS, speeds at the beginning of the route (on W 41st Avenue west of Granville Street and through the UBC Endowment Lands) are higher than further east along the route, where the route passes through denser and more congested segments of W 41st Avenue.

```{r, message = FALSE}
# Avg speed R4 EB 8AM
avg_speed_8am <- 
  r4_trips_ph_speed |> filter(hour_dep == 8) |> pull(speed)

stop_seq_speeds_8am <- 
translink$stop_seq |> 
  filter(itin_id == r4_e_itin_id) |> 
  left_join(translink$stops |> as.data.frame() |> select(stop_id,stop_name),
  by= "stop_id") |> 
    mutate(speed = round(avg_speed_8am*speed_factor,1)) |> 
    select(stop_sequence,stop_name,speed)

print(stop_seq_speeds_8am)

# Visualise trips per hour and speeds in a ggplot
ggplot2::ggplot(stop_seq_speeds_8am[1:nrow(stop_seq_speeds_8am)-1,],aes(x = stop_sequence))+
  geom_line(aes(y = speed)) +
  scale_y_continuous(
    name = "Speed (km/h)")+
  scale_x_continuous(breaks = stop_seq_speeds_8am$stop_sequence)+
  labs(
    title = "R4 Eastbound: 8AM weekday speeds by departure_stop",
    x = "Stop sequence") +
  theme_minimal()
```
